---
title: "lm-ggplot-lab"
author: "Amy Willis, Sarah Teichman and Maria Valdez"
date: "2024-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this lab, we'll practice running linear regression with the function `lm`. 
We'll also manipulate our data with the functions from the `tidyverse` and plot our data with `ggplot`.

WARNING!!: There will be a few intentional errors here for you to fix (so you can practice debugging!). If a line of code isn't working, think about how you can fix it to make it work. 

## Loading Data

Let's begin by loading in the tidyverse, which we'll need for this tutorial.

```{r}
library(tidyverse)
```

Next, let's load our data. We'll start with two data sets. This data comes from a study of people living with cystic fibrosis. It consists of 16S and droplet digital PCR (ddPCR) data from both saliva and sputum (coughed up mucus) from study participants. More information can be found here: 
https://journals.asm.org/doi/full/10.1128/mSystems.00296-20?rfr_dat=cr_pub++0pubmed&url_ver=Z39.88-2003&rfr_id=ori%3Arid%3Acrossref.org&

```{r}
ddpcr <- read_csv("https://raw.githubusercontent.com/statdivlab/stamps2024/main/stats-labs/lm-lab/data/ddPCR.csv")
meta <- read_csv("https://raw.githubusercontent.com/statdivlab/stamps2024/main/stats-labs/lm-lab/data/meta_data.csv")
```

## Preparing the data

We can take a look at the beginning of each dataset with the function head

```{r}
head(ddpcr) #Correction here for the 
head(meta)
```

We'd like to combine these two data sets together. They both contain the variable `sample_name`, so we will match observations based on this variable. `inner_join` combines these two datasets into a single one, cleverly using the `by` argument to figure out how to rows match up. 

Note that when we use `tidyverse` syntax, the pipe symbol %>% is our friend. The pipe lets us chain commands together, using the output of the previous command as the input to the next command. 

```{r}
both <- meta %>% 
  inner_join(ddpcr, by = "sample_name") 
```

Next, we will manipulate our data to make it easier to work with. We'll rename the variable `Average Replicates, stock DNA (copies /µl)` (remember, we like variable names to be concise and avoid spaces) and tell R that this variable is a number(removing the commas in the process). 

We'll define FEV1 to be a numeric variable as well.

```{r}
both <- both %>%
  rename(ddpcr = `Average Replicates, stock DNA (copies /µl)`) %>%
  mutate(ddpcr = gsub(",", "", ddpcr)) %>%
  mutate(ddpcr = as.numeric(ddpcr)) %>%
  mutate(FEV1 = as.numeric(FEV1)) 
```
The last line guarantees `Subject ID` (which is originally a string variable ordered by alphabetical order), is a categorical variable sorted by its numerical value. 

**Optional exercise:** Later in this lab, we will visualize our data using various plots where Subject ID is utilized as a color variable. Subject ID is a categorical variable that can only take a value from a finite set of options (1, 2, 3, ..., 10). However, it is currently sorted alphabetically, causing ID 10 to appear before 2, 3, ..., 9. To correct this, you can modify the variable in the table by converting it to numeric for proper sorting (using `as.numeric`) and then back to a categorical variable (using `factor`). Complete the code below to perform this modification before proceeding to the plotting section. 

```{r}
both <- both %>%
  mutate(`Subject ID` = factor(as.numeric(`Subject ID`))) # Completed and corrected name of variable
```

You can run each of these lines individually to further explore what each command is doing. 

## Exploring the Data

Earlier we looked at the first few observations of each dataset. To see a full list of variables in our dataset, we can use the `names()` function.

```{r}
names(both)
```

We'll be focusing on a few variables from this dataset: ddpcr, FEV1, Treatment Group, Sample Type, and Subject ID. 

ddpcr: droplet digital PCR, giving the observed total bacterial load in the sample in copies/uL
FEV1: Forced expiratory volume (FEV) measures how much air a person can exhale during a forced breath (L/s)
Treatment Group: "ON" if the sample is from a subject receiving antibiotics, "OFF" otherwise 
Sample Type: the body site of origin of the sample, either Saliva or Sputum
Subject ID: an ID identifying the participant 

Before we do inference, we should plot our data. This lets us get a sense of general trends and build intuition about our data. 

We'll use ggplot for plotting. Remember that in ggplot we start with the base of our plot, where we specify our dataset and variables, and then we can build up our plot by adding layers with the `+` operator. We'll build this first plot step by step.

```{r}
ggplot(data = both, aes(x = `Sample Type`, y = ddpcr, col = `Subject ID`))
```

This sets up our axis labels. We are now ready to add a plotting layer.

```{r}
ggplot(data = both, aes(x = `Sample Type`, y = ddpcr, col = `Subject ID`)) +
  geom_jitter(height = 0, width = 0.3)
```

geom_jitter is a great way to make a scatterplot with a categorical variable on the x-axis. It spreads the points out so that they do not all line up exactly for the same values of x. You can change the width value to see how it changes the appearance of the plot.

We like to add informative labels with the `labs` layer to improve on our variable names. Select proper names for the y-axis and the title of the plot below (don't forget to use quotation marks!):

```{r}
ggplot(data = both, aes(x = `Sample Type`, y = ddpcr, col = `Subject ID`)) +
  geom_jitter(height = 0, width = 0.3) + 
  labs(y = "Average observed ddPCR\n(copies/ul)",
       title = "Digital PCR by Treatment") #Added the labels here
```


ggplot by default left-justifies titles. The theme layer centers the title. Add `theme(plot.title = element_text(hjust = 0.5))` to improve the look of your plot above.

The ddpcr data is skewed, with a few larger outliers taking up a lot of space on our plot. We can transform ddpcr to the log of ddpcr to visualize this in a  different way.

```{r}
ggplot(data = both, aes(x = `Sample Type`, y = ddpcr, col = `Subject ID`)) +
  geom_jitter(height = 0, width = 0.3) + 
  labs(y = "Average observed ddPCR\n(copies/ul)",
       title = "Digital PCR by Treatment") + 
  scale_y_log10()
```

What do we learn from this plot? 
It seems that there are sputum samples generally have higher concentration (ddPCR) values than saliva samples.

```{r}
ggplot(data = both, aes(x = `Treatment Group`, y = ddpcr, col = `Subject ID`))  + #Added the missing + sign
  geom_jitter(height = 0, width = 0.3) +
  labs(y = "Average observed ddPCR\n(copies/ul)",
       title = "Digital PCR by Treatment") + 
  scale_y_log10() + 
  theme(plot.title = element_text(hjust = 0.5))
```

What do we learn from this plot? 
There are a lot more non-treated samples than treated samples. The general trend seems to be higher observed concentration values from non-treated samples than treated samples.

```{r}
ggplot(data = both, aes(x = as.numeric(FEV1), y = ddpcr, col = `Subject ID`)) +
  geom_point() + 
  labs(x = "FEV Value",
       y = "Average observed ddPCR\n(copies/ul)",
       title = "Digital PCR by FEV Value") + 
  scale_y_log10() +
  theme(plot.title = element_text(hjust = 0.5))
```

What do we learn from this? 

There might be a slight negative correlation between concentration values and FEV values
There seem to be two observations for each value on the x-axis. Why might that be?

It turns out, saliva and sputum samples are paired (one each for each subject), and so share a value of FEV1! Whoops - we forgot about this - or maybe the postdoc left and we weren't told. 

Insights like this from our plot might lead us to analyze our data. 

## Fitting regression models

Now, let's look at the association between digital PCR measurements and whether the sample is from saliva or sputum. We'll use the lm command. 
  
Note that we are deciding to model the mean of our outcome on the linear scale. We could also choose to model the log of the mean of this variable. We will see an example of this at the end of this lab. Both options are valid, as long as you interpret the coefficients in terms any transformations you've done. This decision is driven by which model you believe to reflect reality better. 

We use `lm` to fit the linear model. 

```{r}
mod_type <- lm(ddpcr ~ `Sample Type`, data = both)
summary(mod_type)
```

Notice the syntax we use here. The first argument is called a formula. It takes the form "response variable ~ predictor variable". The second argument specifies the dataset that contains the variables you are using. 

We use ` ` around the variable name `Sample Type` because it has a space in it. This is another good reason not to have spaces in your variable names! 

To look at the results of our model, we can use the function `summary`

How do we interpret the previous output? 

We have two rows in the "Coefficients" table. The first row corresponds with the intercept of our model. The following rows correspond with covariates. 

We have 4 columns. The first column gives the estimate of the coefficient for the intercept or covariate on the row. The second column gives the standard error (remember, this is an estimate of the standard deviation of our estimate). 

`lm` will by default run a hypothesis test with the hypothesis that the coefficient for each covariate is equal to 0. If this hypothesis is rejected, we determine that there is statistically significant relationship between our covariate and outcome. The third column of our output gives the test statistic for that hypothesis test, and the fourth column gives a p-value. 

What does the second row of the output tell us? 

Our estimate of 1,006,955 for the coefficient for `Sample Type`Sputum tells us that the estimated mean observed ddPCR reading for a sample of sputum is 1,000,955 counts/uL higher than the estimated mean observed ddPCR for a sample from saliva. 

We have a p-value of 0.01355. If we use a alpha level of 0.05, we reject the null hypothesis that there is no difference in the true mean concentration in Sputum and Saliva samples. 

Now, let's fit a model with multiple covariates. Let's consider `Treatment Group` and `Sample Type`. Write the proper model below, using `+` to connect covariates

```{r}
mod_treat_type <- lm(ddpcr ~ `Sample Type` + `Treatment Group`, data = both) #Completed formula
summary(mod_treat_type)
```

How would we interpret these results? 
Here, we have a significant difference in mean ddPCR across Sample Types (at the alpha=0.05 level), but not across Treatment groups. 

What if we think there is an interaction between the effects of Treatment and Sample Type on ddpcr? i.e., a different change between Treatment ON and OFF for Sputum vs Saliva. 

We can fit an interaction model by replacing the `+` above with `*`.

```{r}
mod_interact <- lm(ddpcr ~ `Sample Type` * `Treatment Group`, both) # Corrected typos and orders of arguments
summary(mod_interact)
```

Now we have an extra row. Not only do we have rows for `Sample Type`Sputum and Treatment, but we also have an interaction row. The estimate for the interaction can be interpreted as the following: 

-349,375 is the estimated value of the difference in mean bacterial concentration in sputum for people on antibiotics compared to people not on antibiotics minus the difference in mean bacterial concentration in saliva for people on antibiotics compared to people not on antibiotics.

Note that our estimates and standard errors for `Sample Type`Sputum and `Treatment Group`ON are different for our two models. This is because whenever we change something (add a covariate, add an interaction, etc) we are fitting a different model. Our strongest advice on the subject of "which model to choose?" is to be motivated by your scientific question, and consider the classification of variables presented in the lecture: predictor of interest (always include), precision variables (good to include most important ones), confounders (always include, but use the actual definition, not the way most people talk about it) and relevant effect modifiers (include if scientifically relevant)). 

Finally, we could model the ddpcr as counts coming from a random process and we are interested in exploring how the mean of these counts may be related to the `Sample Type`. An option for this is using a Generalized Linear Model with a log link function (using a Poisson regression). We can do this in R through `glm`.  

```{r}
mod_counts_type <- glm(ddpcr ~ `Sample Type`, data = both, family = poisson(link = "log"))
summary(mod_counts_type)
```

How do we interpret these coefficients? 

Since in this model we are using a log link function, our coefficients describe how the log(expected counts) differs between saliva samples and sputum samples. A p-value lower than 2e-16 (at an alpha level of 0.05) indicates we have enough evidence to reject there is no difference between these log(expected counts). And by exponentiating  the estimated coefficient (exp(7.336e-01) = 2.082564), we can say that the estimated expected counts observed for samples of Sputum is 2.08 times higher than the estimated expected counts for samples of saliva.

Now create your own model, by selecting:
1. Two covariates that you think could potentially be associated to the ddpcr counts in some scale. 
2. How are you interested in exploring these associations? 
Include your fitted model here: 

```{r}
  My_own_mod <- 
  
  summary(My_own_mod)
```

And then provide an interpretation on your results: 