---
title: "04-dplyr key"
author: "Maria Valdez C. and Sarah Teichman"
date: "2024-07-15"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this tutorial, we will go through how to manipulate data with the `dplyr` package. In previous tutorials, you learned how to load data into R as a data frame object. This tutorial focuses on modifying the data frames in order to clean and reorganize our data before analysis. 

## Lesson 4 - dplyr and Tidyverse

The `tidyverse` is a suite of powerful and cohesive R packages. The `dplyr` package is part of this suite and has commands for common data manipulation. In this tutorial, we will overview the basics of using `dplyr`. 

## Installation 

You can load the whole `tidyverse` suite like this,

```{r}
if (!requireNamespace("tidyverse", quietly = TRUE)) {
     install.packages("tidyverse")
}
library(tidyverse)
```

but optionally, you can individually install and load `dplyr` (or any other package in the suite such as `ggplot`).

```{r}
if (!requireNamespace("dplyr", quietly = TRUE)) {
     install.packages("dplyr")
}
library(dplyr)
```

## dplyr commands

Some of the `dplyr` commands designed for data manipulation operations are: 

`select()` retains columns (typically variables) based on their names.
`filter()` retains rows (typically observations) based on their values.
`arrange()` changes the order of the rows.
`relocate()` changes the order of columns.
`mutate()` adds new columns that are functions of existing columns.
`group_by()` groups all rows that have the same value for a chosen column. 
`summarise()` reduces multiple values down to a single summary.

## Loading data 

We are going to load the data set “phylum_data.txt”, which combines microbial taxon counts and covariates from water samples taken around Woods Hole (it combines `covariates` and some of the rows in `otus` data sets from previous tutorials). 

Recall that we can load data using `read.csv`: 

```{r}
phylum_data <- read.csv("data/phylum_data.txt", sep = "\t", row.names = 1, header = T)
```

## select and filter commands

Sometimes you may want to subset columns (variables) and rows (observations) in your data for an analysis. `select` retains columns based on names, and `filter` retains rows based on values. For example:

```{r}
select(phylum_data, Location, Month, Acidobacteria, Chlamydiae)
```

You can use select with `contains()` or `starts_with()` with `select` if appropriate. `contains()` will include all column names that contain a substring, while `starts_with()` will include all column names that start with the given substring. For example:

```{r}
select(phylum_data, Location, Month, contains("bacteria"))
```

`filter` will subset the data frame by keeping all rows for which a logical expression is true. For example:

```{r}
filter(phylum_data, Month == "Jan")
Jan_phylum_data <- filter(phylum_data, Month == "Jan")
```

You can also `filter` based on multiple criteria. In this case, `|` is the logical "or" and `&` is the logical "and".

```{r}
filter(phylum_data, Month == "Jan" | Month == "Feb")
filter(phylum_data, Season == "Winter" & Location == "MBL")
```

## Exercise

Build a new data frame containing variables "Location", "Season", and only taxa whose names start with "Ch", while preserving only the observations with type "Outfall".

```{r}
Temporary_phylum_data <- filter(phylum_data, Type == "Outfall")
Ch_phylum_data <- select(Temporary_phylum_data, Location, Season, starts_with("Ch"))
Ch_phylum_data
```

*Tip*: Since "Type" is not in the variables to be kept, you should perform `filter` before `select` here, or you'll need to use `select` twice.

## Piping!

It is often tedious to save intermediate data frames while performing multiple data manipulations, like `select` and `filter`. Using the pipe operator, `%>%`, you can chain together multiple commands, with the output from one automatically being used as the input to the next. You can start with the first input (often the data frame), and then pipe multiple functions together:

```{r}
phylum_data %>% 
  select(Location, Month, Acidobacteria, Chlamydiae) # using a single function 

Jan_phylum_data <- phylum_data %>% 
  select(Location, Month, Acidobacteria, Chlamydiae) %>%
  filter(Month == "Jan") # using multiple functions

Jan_phylum_data
```

Repeat the previous exercise with piping (does the order of operations still matter?)!

**note:** the following code should fail to run
```{r}
#The order of operations does matter. This code will fail: 
Ch_phylum_data <- phylum_data %>% 
  select(Location, Season, starts_with("Ch")) %>%
  filter(Type == "Outfall")
```

```{r}
#The order of operations does matter. This code will fail: 
Ch_phylum_data <- phylum_data %>% 
  filter(Type == "Outfall") %>%
  select(Location, Season, starts_with("Ch"))

Ch_phylum_data
```

## rename and relocate

The last column in our data set is called "phylum_NA", which represents counts for taxa that we do not recognize. We could change the name to a more descriptive name such as "unknown". We do this with the `rename()` function. We'll use `head()` to avoid printing out the entire dataset.

```{r}
phylum_data %>%
  rename(unknown = phylum_NA) %>%
  head()
```

We can also `relocate` columns. For example, if we are interested in easily comparing the counts of unknown organisms against the covariates in our exploration, we could use:

```{r}
phylum_data %>% 
  rename(unknown = phylum_NA) %>%
  relocate(unknown, .after = Type) %>%
  head()
```

In `relocate`, `.before` indicates "the columns should come before ...", and `.after` indicates "the columns should come after ...".

## mutate & arrange

`mutate()` is used to create new columns. Imagine that we want the log counts of "Actinobacteria". We can create a modified data frame with a new column named `log_Actinobateria` (we are also using relocate to put `log_Actinobacteria` right after `Actinobacteria`):

```{r}
phylum_data %>%
  mutate(log_Actinobacteria = log(Actinobacteria)) %>% 
  relocate(log_Actinobacteria, .after = Actinobacteria) %>%
  head()
```

`arrange()` is used to order the rows of the dataset depending on the values of one or more variables.
Let's order `phylum_data` based on the counts of *Actinobacteria* observed. By default this will sort from small to large, or you can use `desc` to reverse the order.

```{r}
phylum_data %>%
  arrange(Actinobacteria) %>%
  head()
```

```{r}
phylum_data %>%
  arrange(desc(Actinobacteria)) %>%
  head()
```

## Exercise

With all you have learned, create a new data frame in which you maintain only the covariates and the *Acidobacteria* column. For this taxon, preserve only the positive counts (discard zero values) and obtain the log counts. Order samples in decreasing order of these counts. Do you notice any trend based on the season?

```{r}
Acidobacteria_data <- phylum_data %>%
  select(Location, Month, Season, Type, Acidobacteria) %>%
  filter(Acidobacteria > 0) %>%
  mutate(log_Acidobacteria = log(Acidobacteria)) %>%
  arrange(desc(Acidobacteria))

Acidobacteria_data
```

Here we can see that most samples in which *Acidobacteria* were observed were collected in the winter.

## group by and summarize

`group_by()` allows you to define groups within the data set based on specific variables. Grouping the data is often a precursor to performing summary operations on the grouped data, like `summarize`. 

The `summarize()` function is used to calculate summary statistics or perform other operations on grouped data. It allows you to specify summary functions (e.g., `mean()`, `sum()`, `max()` or custom functions) to apply to a specific variable for each group. The result is a new summarized data set where each row represents a group and each column contains the summarized values.

Returning to our example, say instead we want to get total counts of *Acidobacteria* in winter vs in summer. We can obtain this by:

```{r}
phylum_data %>%
  group_by(Season) %>% 
  summarize(Acidobacteria_tot = sum(Acidobacteria))
```

Note that `group_by()` can include multiple variables.

```{r}
phylum_data %>%
  group_by(Season, Type) %>% 
  summarize(Acidobacteria_tot = sum(Acidobacteria))
```

## Challenge exercise 

Create a new data frame, finding the log counts of "Chlamydiae" by "Location" and "Season", removing zero counts.

```{r}
new_data <- phylum_data %>%
  group_by(Location, Season) %>%
  summarize(Chlamydiae_tot = sum(Chlamydiae)) %>%
  filter(Chlamydiae_tot > 0) %>%
  mutate(log_Chlamydiae_tot = log(Chlamydiae_tot))

new_data
```
