---
title: '05: pivots and plotting with ggplot key'
author: "Shirley Mathur and Sarah Teichman"
date: "2024-07-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, let's load the `tidyverse` packages and read in our data. For this tutorial, we will be working with the `phylum_data` dataset, which includes metadata and microbial taxon counts for samples taken around Woods Hole.

```{r, message=FALSE}
library(tidyverse)
```

```{r}
phylum_data <- read_tsv("data/phylum_data.txt")
```

## Pivots

Pivoting lets you reorganize the structure of your data by rearranging columns and rows. This can be crutial to getting data in the right format for plotting (you might be familiar with this concept from excel, or `melt` in python).

You can pivot data wide or long. Wide data has more columns, and there is some type of data stored in the column names (e.g. sample, year, location). Long data has fewer columns, and the data that used to be part of the column names is now stored as values in a new column.

The `pivot_wider()` function in R is used to transform data from a longer format to a wider format, allowing you to spread values across multiple columns based on a key-value pair. Conversely, the `pivot_longer()` function is used to transform data from a wider format to a longer format, where multiple columns are combined into key-value pairs, providing a more compact representation of the data.

Let's try this with our table. Recall the column names:

```{r}
names(phylum_data)
```

Notice that many of the column names are taxon names. We can pivot the data longer, by creating a column called "taxon", and storing the count data in a "count" column.

```{r}
phylum_data_long <- phylum_data %>% 
  pivot_longer(
    !c(Sample, Location, Month, Season, Type), #pivot all the rows except for the sample metadata
    names_to = "taxon", values_to = "count"
  )
phylum_data_long %>% head()
```

We can recover our original table by pivoting wider by taxon, as shown below.

```{r}
phylum_data_long %>% pivot_wider(names_from = taxon, values_from = count) %>% head()
```

## Into to Plotting 

Now, we will delve deeper into plotting with `ggplot2`.
 
`ggplot2` (commonly just referred to as just ggplot) is a R package used to make high quality plots for your data. *ggplot2 is part of the tidyverse suite!* It offers consistent syntax and allows for detailed customization of almost every element of your plots. Visualization options range from bar graphs, boxplots, scatterplots, histograms and more. 

First, we will summarise the total counts for each taxon across all of our samples using the functions we learned from the `dplyr` tutorial.

```{r}
phylum_summary <- phylum_data_long %>%
  group_by(taxon, Location, Season) %>%
  summarise(count = sum(count))

phylum_summary %>% head()
```

Now, let's take our summarized taxon counts data frame and make a plot!

First, let's look up the help documentation for `ggplot` to see what information
it needs from us to make a plot.

```{r}
help(ggplot)
```

The two required arguments are `data` and `mapping`.

We will start by using the `ggplot`function to make a plot, and setting the `data` argument to be `phylum_summary`, and then set the `mapping` argument to be `aes(x = taxon, y = count)` to give the aesthetics of our plot, specifying that we want the taxon labels on the x-axis and their total counts on the y-axis.

The below code will create a plot that has no data on it, but just the aesthetics that we specified.

```{r}
ggplot(phylum_summary, aes(x = taxon, y = count)) 
```

Using piping, this is equivalent to:

```{r}
phylum_summary %>% ggplot(aes(x = taxon, y = count)) 
```

Now, say we want to make a scatter plot of our taxon counts. This is about the simplest plot we can make with `ggplot`.

To do this, we will add a layer to our plot for a scatter plot, which is the
`geom_point()` layer.

```{r}
ggplot(phylum_summary, aes(x = taxon, y = count)) + 
  geom_point()
```

However, this plot is perhaps not the best way to show our data, as we are interested in plotting a summary of the counts. We could try using a bar plot instead.

For a bar plot, we will now add a `geom_col()` layer instead of the point layer.

```{r}
ggplot(phylum_summary, aes(x = taxon, y = count)) +
  geom_col()
```

We can note that the above code is equivalent to:

```{r}
phylum_summary %>% ggplot(aes(x = taxon, y = count)) +
  geom_col()
```

We will typically see the second code, in which we pipe our data in `ggplot` used more often. We will demonstrate below the utility of this approach.

Comparing the bar plot to the scatter plot, you may wonder why in the scatter plot we had 6 different points for each taxon and in the barplot we only have one bar for each taxon. This is because the bar plot internally summarizes the data (summing taxon counts across samples) to make the bars.

This is equivalent to if we have used our `dplyr` operations to summarize the data, providing a single row per taxon:

```{r}
phylum_summary %>%
  group_by(taxon) %>%
  summarise(count = sum(count)) %>%
  ggplot(aes(x = taxon, y = count)) +
  geom_col()
```

We can see that it is the same as our other bar plot, which did the summarising internally.
 
Now, the x-axis above is a bit muddled and difficult to read. Let's clean this up a little bit - we can fix those axis labels like this:

```{r}
phylum_summary %>% 
  ggplot(aes(x = taxon, y = count)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #hjust sets the alignment of the text.
  geom_col()
```

Now, we can add meaningful axes labels and a title by adding a `labs` layer that specifies what labels we want to add to our plot for the axes.

```{r}
phylum_summary %>% 
  ggplot(aes(x = taxon, y = count)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(title = "Taxon counts", x = "Taxon", y = "Count")
```

Now, we'll add another aesthetic, `fill`, which will let us fill in the bar color by a variable of interest. Here, we can consider subsetting counts based on location.

```{r}
phylum_summary %>% 
  ggplot(aes(x = taxon, y = count, fill = Location)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(title = "Taxon counts", x = "Taxon", y = "Count")
```

In this plot each of the bars are segmented by color, where the color is showing how much of a particular taxon's counts come from each location.

In practice however, it might not be particularly useful to just have a color-coded segmented bar chart as shown above. Instead, we might want to create plots for how the counts vary across different values of the season and location covariates.

We can accomplish this by adding a `facet` layer to our plot, and specifying which variables we want to facet across. Here we can facet by "Location".

```{r}
phylum_summary %>% 
  filter(Season == "Winter") %>%
  ggplot(aes(x = taxon, y = count, fill = Location)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~ Location) +
  labs(title = "Taxon counts", x = "Taxon", y = "Count")
```

In the plot above, we can now see how the total taxon counts vary across the different locations.

Now, it is time for you to practice!

## Exercise

Fill in the code below to make a facet plot that plots the counts across different combinations of seasons and locations, and so that you have the different locations across the rows and different seasons across the columns.

(Hint: edit the facet formula to facet across seasons and location)

*Note: here, we are now using the `facet_grid` function rather than `facet_wrap`. This function is more useful when you want to facet across combinations of two variables, whereas `facet_wrap` is better when you have one variable to facet across and you want to specify how many rows, columns you want in your faceted plot.


```{r, eval=FALSE}
phylum_summary %>% 
  ggplot(aes(x = taxon, y = count, fill = Location)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_grid(Location ~ Season) +
  labs(title = "Counts of Taxa", x = "Taxon", y = "Count")
```

## Exercise

Earlier, we were looking at raw counts across different covariate values and plotting them using bar plots. 

However, we might be interested in how different different taxon's counts change together. In particular, let's say we are interested in seeing how *Cyanobacteria* and *Proteobacteria* counts vary together. This is a setting where a scatterplot would be more useful.

Use your `ggplot` skills to make this scatter plot using the `phylum_data` table.

*Hint*: pipe the data into `ggplot`, set appropriate x and y aesthetics, and add an appropriate plotting layer to make a scatter plot.

```{r}
phylum_data %>% ggplot(aes(x = Proteobacteria, y = Cyanobacteria)) +
  geom_point()
```
